% 00-abstract.tex — 中英文摘要

\begin{center}
{\large\bfseries\color{Vermilion} 摘\quad 要}
\end{center}

\noindent
针对元胞自动机是否生成迷宫这一判定难题,本文给出两件互补的产物。

其一,$\textbf{FamilyMask}$ 多族染色体编码。它把经典 Conway 型 B/S 规则推广为多族
组合:每族独立持有自己的邻域 mask (固定 $9\times 9$ 中心外 80 格)、
出生集、存活集与优先级;最多 16 族并行,执行时按优先级顺序遍历,首个匹配的族
决定下一态。单族时退化为标准 B/S 字符串 ($2^{18}$ 候选),多族时搜索空间跳跃至
一族 B/S 加一组 mask 的 $2^{103}$ 候选级 (1 active + 80 cells + 9 B + 9 S + 4 priority = 103 bits,见 \texttt{src/search/rule\_chromosome.js::SLOT\_BITS})。

其二,$\textbf{maze\_quality}$ 八维几何质量度量。它在 $\min(M_{\text{topology}},
\,M_{\text{diversity}}) \times M_{\text{wall\_gate}}$ 的两层结构上聚合八个连续可微
子度量,所有取值 $[0,1]$,ES 可学。具体而言,五个拓扑子度量
(走廊主导度 $M_b$、最长路径占比 $M_s$、十字口占比 $M_j$、最大连通分量占比 $M_c$、
外圈抑制 $M_{\text{bnd}}$) 加权几何聚合得到 $M_{\text{topology}}$;三个多样性
子度量 ($2\times 2$ 块熵 $M_p$、镜像不对称度 $M_a$、$2\times 2$ 邻接对熵
$M_t$) 加权几何聚合得到 $M_{\text{diversity}}$。再用 $\min$ 强制两侧平衡,乘
以 $[0.40,\,0.60]$ 上值为 1 的墙比软三角门控 $M_{\text{wall\_gate}}$,
拦截 ``实心块'' 与 ``空盒'' 两种典型病态吸引子。

在 15-pattern 对照基准 (六个经典迷宫生成算法 DFS/Kruskal/Prim/Growing Tree/
Sidewinder/Binary Tree + 九个非迷宫吸引子 Spiral/Fractal Tree/Horizontal Stripes/
Random Noise (50%/30%)/Checkerboard/Diagonal Stripes/Concentric Rings/Honeycomb)
上,$maze\_quality$ 真正迷宫均值 $0.711$,伪迷宫均值 $0.000$,间隔 $+0.711$,
9 个伪迷宫无误判;同一数据集上 Bellot 2021 $F = \nu/\delta$ 度量真伪间隔 $-191.3$,
且 9 个伪迷宫中有 4 个 ($F = 0$ 的 Horizontal Stripes/Concentric Rings/Honeycomb
加上 $F = 11.85$ 的 Spiral) 排名反低于真正的 DFS ($F = 15.33$) ,
即 Bellot F 错误判定 ``Spiral 比 DFS 更像迷宫'',这是
附录~D 系统审计 (\texttt{tab:repro-audit}) 中已经指出的 smoking gun
(本节下方全部使用 \texttt{src/gpu/bellot\_metrics.js::bellotF} 真实 Bellot 2021 实现
$F = \nu(M)/\delta(M)$,不再使用早期 \texttt{scripts/full\_mq\_v2.mjs} 的简化版
$\nu = 1-\text{roadFrac},\, \delta = d/200$)。

将 FamilyMask 嵌入 $200\!\times\! 500$ 的 $(\mu{+}\lambda)$ 进化策略,WebGPU
compute shader 加速 CA 评估,在 128 次 sweep (8 距离模板 $\times$ 4 族数
上限 $\times$ 4 随机种子, $40\times 60$ 网格, 300 步演化) 上,成功完成 122 次,
得到最高分 $\mathbf{0.8233}$ (manhattan-2 模板、8 族上限、种子 444)。进一步
统计得到三组可重复观察:(i) 多族容量上限升高反而使均值下降 (mf=1: 0.6984
$\to$ mf=8: 0.6306),但 mf=8 的尾部拥有 0.8233 全场最高,体现 ``均值降 /
峰值升'' 的 ``偶发突破'' 现象;(ii) chebyshev-4 模板均值仅 0.157,主因为其
$9\times 9$ 邻域使单族搜索空间维度达 $2^{103}$,远大于 $10^5$ 次评估预算可跨越
的初始化-最优距;(iii) manhattan-1 (4 邻域 von Neumann) 模板均值仅 0.420,
原因是 4 邻域无法为走廊结构提供足够的二阶空间上下文。所有代码、实验数据、
截图开源。

\vspace{0.6em}
\begin{center}
{\large\bfseries\color{Vermilion} Abstract}
\end{center}

\noindent
We present two complementary contributions for evaluating whether binary
grids generated by cellular automata (CA) constitute mazes.
\textbf{FamilyMask} generalises Conway-style life rules to a multi-family
chromosome: each of up to sixteen families owns its own neighbourhood mask
(the $9\times 9$ centred footprint minus the centre, 80 cells), birth set,
survival set, and priority; at each step the CA scans active families in
priority order and the first match decides the next cell state.
\textbf{maze\_quality} is an eight-dimensional geometric score aggregating
five topology sub-metrics (corridor dominance $M_b$, longest-path share
$M_s$, junction share $M_j$, largest-component share $M_c$, outer-frame
suppression $M_{\text{bnd}}$) and three diversity sub-metrics
($2\times 2$ patch entropy $M_p$, mirror asymmetry $M_a$, $2\times 2$
adjacency-pair entropy $M_t$) via a two-level weighted geometric mean,
clamped by $\min$ and multiplied by a triangular wall-ratio gate centred at
$[0.40,\,0.60]$. On a 15-pattern benchmark the score attains mean $0.711$
on true mazes and $0.000$ on pseudo-mazes (gap $+0.711$, zero
misclassifications of 9 pseudo-patterns); the Bellot 2021 $F = \nu/\delta$
score on the same dataset has gap $-191.3$, with the wrong sign
(``lower=maze-like''), and 4 of 9 pseudo-mazes (Spiral $F{=}11.85$ plus
three $F{=}0$ strip/ring/honeycomb patterns) are ranked BELOW the genuine
DFS ($F=15.33$) --- the smoking gun.

Plugging FamilyMask into a $200$-population, $500$-generation
$(\mu{+}\lambda)$ evolution strategy and running 128 ES sweeps over
8 distance templates, 4 family caps, and 4 random seeds ($40\times 60$
grids, 300 CA steps, WebGPU-accelerated) yields 122 successful runs and a top
score of $\mathbf{0.8233}$ (manhattan-2, 8-family cap, seed 444). On the
15-pattern benchmark the score attains mean $0.711$ on true mazes and $0.000$
on pseudo-mazes (gap $+0.711$, zero misclassifications of 9 pseudo-patterns);
the Bellot 2021 $F = \nu/\delta$ score on the same dataset has gap $-191.3$
(``lower=maze-like''), and 4 of 9 pseudo-mazes (Spiral $F=11.85$ plus three
$F{=}0$ strip/ring/honeycomb patterns) are ranked BELOW the genuine DFS
($F = 15.33$) --- the smoking gun. Three robust observations emerge: (i)~larger family caps lower the mean
(mf=1: 0.6984 $\to$ mf=8: 0.6306) but raise the upper tail (mf=8 owns the
overall best 0.8233); (ii)~chebyshev-4's poor mean ($0.157$) is caused by the
the $9\times 9$ neighbourhood implying a $2^{103}$ candidate space (per family) far beyond
the $\sim\!10^5$ evaluations available to random-init-to-optimum distance;
(iii)~manhattan-1's poor mean ($0.420$) is caused by the 4-neighbour
(von~Neumann) topology providing too little spatial context for corridor
emergence. All code, data, and figures are released as open source.
