% 00-abstract.tex — 中英文摘要

\begin{center}
{\large\bfseries\textcolor{Vermilion}{摘\quad 要}}
\end{center}

\noindent
本文解决一个具体的判定问题: 给定一个二值网格 (元胞自动机运行结果), 如何判断它是否构成一个迷宫? 这个判断并非一目了然, 因为 Conway 生命游戏的演化常常产生 ``看起来像迷宫, 实际不是'' 的吸引子 (实心块、空盒、规则条纹、随机噪声), 而 ``真迷宫'' 又有走廊结构、稀疏分叉、外圈封闭等拓扑约束。

\noindent\textbf{FamilyMask 多族 CA 编码。}把经典 B/S 型规则推广为最多 16 族并行、按优先级仲裁的染色体: 每族独立持有自己的 $9 \times 9$ 邻域 mask (去掉中心 80 格)、出生集、存活集与优先级; CA 每步按优先级遍历 active 族, 首个匹配的族决定下一态。单族时退化为标准 B/S ($2^{18}$ 候选), 多族时搜索空间跃升至 $2^{103}$ 单族 / $2^{1648}$ 全染色体级 (16 slot $\times$ 103 bits/slot, 103 = 1 active + 80 cells + 9 B + 9 S + 4 priority)。

\noindent\textbf{maze\_quality 八维几何度量。}在 $\min(M_{\text{topology}}, M_{\text{diversity}}) \cdot M_{\text{wall\_gate}}$ 的两层结构上聚合 8 个连续可微子度量: 5 个拓扑 ($M_b$ 走廊主导度、$M_s$ 最长路径占比、$M_j$ 十字口占比、$M_c$ 最大连通占比、$M_{\text{bnd}}$ 外圈抑制) + 3 个多样性 ($M_p$ $2{\times}2$ 块熵、$M_a$ 镜像不对称度、$M_t$ $2{\times}2$ 邻接对熵)。所有取值 $[0,1]$, ES 可学; $\min$ 而非几何均值强制两侧平衡, 三角墙比门控 $M_{\text{wall\_gate}}$ 拦截 ``实心块 / 空盒'' 病态吸引子。\textbf{v2.1 权重}: 5 个拓扑子度量加权几何聚合采用 $0.20/0.15/0.20/0.30/0.15$ ($M_b/M_s/M_j/M_c/M_{\text{bnd}}$), 是经过多轮实验、结合人眼评判迭代调优的结果, 让 ``分数排名与视觉评判一致''。$M_c$ 改为纯线性 (旧版有 $/0.8$ clip 导致 $M_c \in [0.8, 1.0]$ 全部饱和到 1.0); 旧版权重在 ES 跑出的 4 个真迷宫 ($M_c \in [0.74, 0.98]$) 实际无区分度, 新版下视觉上更接近迷宫形态的 (b) chebyshev-1 ($M_c = 0.98$) 升到榜首, 视觉上走廊感较弱的 (a) manhattan-2 ($M_c = 0.74$) 调整到第三, 完成了 ``分数排名与视觉评判一致'' 的关键修复。

\noindent\textbf{15-pattern 对照基准。}6 个真迷宫生成器 (DFS / Kruskal / Prim / Growing Tree / Sidewinder / Binary Tree) + 9 个几何反例 (Spiral / Fractal Tree / Horizontal Stripes / Random Noise 50\% / Random Noise 30\% / Checkerboard / Diagonal Stripes / Concentric Rings / Honeycomb)。$maze\_quality$ 真迷宫均值 $0.711$, 伪迷宫均值 $0.000$, 间隔 $+0.711$, 误判 $0/9$。同一数据集上 Bellot 2021 $F = \nu/\delta$ 真伪间隔 $-191.3$, 方向反转; 9 个伪迷宫中有 4 个 ($F{=}0$ 的 Horizontal Stripes / Concentric Rings / Honeycomb 加上 $F{=}11.85$ 的 Spiral) 排名低于真迷宫最低值 ($F{=}15.33$, DFS), 即 Bellot $F$ 错误判定 ``Spiral 比 DFS 更像迷宫''。

\noindent\textbf{两个规模的演化扫描。}小范围 sweep: 128 次全交叉 (8 mask $\times$ 4 maxFam $\times$ 4 seed, $40 \times 60$ 网格, pop 200, gens 500), 122 次 OK, 最高分 $0.8233$ (manhattan-2 / maxFam=8 / seed=444)。大范围 sweep: 5 次 (manhattan-2 / maxFam=8 单配置, $500 \times 2000$ 网格, pop 500, gens 2000, 5 随机种子), 5/5 完成, 最高分 $0.8195$ (seed=444), 5-run 均值 $0.8020$。两个 sweep 峰值差 $0.0037$, 表明 manhattan-2 / maxFam=8 在 $40 \times 60$ 网格上已接近该配置的理论上界。

\noindent\textbf{两类失败模式。}(1) chebyshev-4 均值 $0.157$: $9 \times 9$ 邻域使单族搜索空间 $2^{103}$, $10^5$ 预算下 ES 在无梯度平台上原地踏步 --- 预算受限型。(2) manhattan-1 均值 $0.421$: 4 邻域无法为走廊结构提供足够二阶空间上下文 --- 表征受限型。二者分属不同类别, 混淆会误导调参方向。

\vspace{0.5em}
\begin{center}
{\large\bfseries\textcolor{Vermilion}{Abstract}}
\end{center}

\noindent
We study the problem of deciding whether a binary grid produced by a cellular automaton is a maze. Conway-style CA evolution tends to produce many ``looks like a maze, isn't'' attractors (solid blocks, empty boxes, regular stripes, random noise), so the question needs a quantitative answer. Two contributions.

\noindent\textbf{FamilyMask} generalises Conway B/S rules to a multi-family chromosome. Up to 16 families run in parallel; each family owns its $9{\times}9$ neighbourhood mask (the centred footprint minus the centre, 80 cells), birth set, survival set, and 4-bit priority. The CA scans active families in priority order each step; the first match decides the next cell. With one active family the rule space is the standard B/S string ($2^{18}$); with $k$ active families it jumps to $2^{103}$ per family and $2^{1648}$ for the full 16-slot chromosome.

\noindent\textbf{maze\_quality} is an eight-dimensional geometric score. Five topology sub-metrics (corridor dominance $M_b$, longest-path share $M_s$, junction share $M_j$, largest-component share $M_c$, outer-frame suppression $M_{\text{bnd}}$) form $M_{\text{topology}}$ via weighted geometric mean. Three diversity sub-metrics ($2{\times}2$ patch entropy $M_p$, mirror asymmetry $M_a$, $2{\times}2$ adjacency-pair entropy $M_t$) form $M_{\text{diversity}}$. The two sides are combined with $\min$, then multiplied by a triangular wall-ratio gate $M_{\text{wall\_gate}}$ on $[0.40, 0.60]$ that zeroes out solid blocks and empty boxes. All eight sub-metrics are continuous in $[0,1]$, so evolution strategies can use them directly. The $\min$ (rather than geometric mean) is what blocks the failure mode where one side is high and the other is low. \textbf{v2.1 weights}: the five topology weights are $(0.20, 0.15, 0.20, 0.30, 0.15)$ for $(M_b, M_s, M_j, M_c, M_{\text{bnd}})$, refined through multiple experimental rounds with human visual judgment as the ground truth, and $M_c$ is now linear (a previous $/0.8$ clip saturated $M_c \in [0.8, 1.0]$ to 1.0, killing the dimension's gradient). The new weights restore the visual-consistency invariant: chebyshev-1 (visually the most maze-like, $M_c = 0.98$) takes the top rank, while manhattan-2 ($M_c = 0.74$) drops to third, aligning the score ranking with the authors' visual assessment.

\noindent\textbf{15-pattern benchmark.} Six classical maze generators (DFS, Kruskal, Prim, Growing Tree, Sidewinder, Binary Tree) and nine non-maze attractors (Spiral, Fractal Tree, Horizontal Stripes, Random Noise at 30\% and 50\%, Checkerboard, Diagonal Stripes, Concentric Rings, Honeycomb). $maze\_quality$ gives $0.711$ on the true mazes and $0.000$ on the pseudo-mazes, with zero misclassifications of the nine pseudo-patterns. The Bellot 2021 $F = \nu/\delta$ score on the same dataset has gap $-191.3$ and the wrong sign (``lower is more maze-like''), and four of nine pseudo-mazes (Spiral at $F{=}11.85$ plus three $F{=}0$ patterns) rank below the genuine DFS ($F{=}15.33$). That direction reversal is the smoking gun that motivates $maze\_quality$.

\noindent\textbf{Two scales of evolution sweeps.} Small sweep: 128 full-factorial runs (8 masks $\times$ 4 maxFam $\times$ 4 seeds, $40{\times}60$ grid, pop 200, gens 500), 122 OK, top score $0.8233$ (manhattan-2, maxFam=8, seed 444). Large sweep: 5 runs (manhattan-2/maxFam=8 only, $500{\times}2000$ grid, pop 500, gens 2000, 5 random seeds), all 5 complete, top $0.8195$ (seed 444), 5-run mean $0.8020$. The two tops differ by $0.0037$, so the $40{\times}60$ configuration is already near the configuration's theoretical ceiling.

\noindent\textbf{Two failure modes — and a clean diagnostic that distinguishes them.} chebyshev-4 reaches mean $0.157$ because its $9{\times}9$ neighbourhood gives a $2^{103}$ per-family space, and $10^5$ evaluations are too sparse to escape the gradientless plateau (budget-limited failure, solvable by raising budget or shrinking the mask). manhattan-1 reaches $0.421$ because 4-neighbour topology cannot sustain the second-order survival conditions that corridors need (representation-limited failure, requires a different template). Separating these two categories prevents wasted tuning effort in the wrong direction, and is one of the paper's methodological contributions.

\noindent All code, data, figures, and the 15-pattern benchmark dataset are released as open source.
