% 00-abstract.tex — 中英文摘要

\begin{center}
{\large\bfseries\textcolor{Vermilion}{摘\quad 要}}
\end{center}

\noindent
本文研究一个具体的判定问题: 给定一个由元胞自动机 (CA) 演化得到的二值网格, 如何判断其是否构成迷宫。该问题并非一目了然, 因为 Conway 生命游戏类的演化常常产生 ``视觉上类似迷宫, 实质上并非迷宫'' 的吸引子 (实心块、空盒、规则条纹、随机噪声等), 而 ``真迷宫'' 又需满足走廊结构、稀疏分叉与外圈封闭等拓扑约束。

\noindent\textbf{FamilyMask 多族 CA 编码.} 本文将经典 B/S 型规则推广为最多 16 族并行、按优先级仲裁的染色体表示: 每族独立持有 $9 \times 9$ 邻域 mask (去除中心格, 共 80 位)、出生集、存活集与优先级; CA 每步按优先级遍历 active 族, 首个匹配的族决定下一态。单族情形退化为标准 B/S 字符串 ($2^{18}$ 候选); 多族情形下, 单族搜索空间扩展为 $2^{103}$, 全染色体扩展为 $2^{1648}$ (16 slot $\times$ 103 bits/slot, 其中 103 = 1 active + 80 cells + 9 B + 9 S + 4 priority).

\noindent\textbf{maze\_quality 八维几何度量.} 本文提出在 $\min(M_{\text{topology}}, M_{\text{diversity}}) \cdot M_{\text{wall\_gate}}$ 的两层结构上聚合 8 个连续可微子度量: 5 个拓扑子度量 (走廊主导度 $M_b$、最长路径占比 $M_s$、十字口占比 $M_j$、最大连通占比 $M_c$、外圈抑制 $M_{\text{bnd}}$) 与 3 个多样性子度量 ($2 \times 2$ 块熵 $M_p$、镜像不对称度 $M_a$、$2 \times 2$ 邻接对熵 $M_t$). 所有子度量取值 $[0, 1]$, ES 可直接优化; $\min$ 聚合 (而非几何均值) 强制两侧平衡, 三角墙比门控 $M_{\text{wall\_gate}}$ 在 $[0.40, 0.60]$ 区间外线性衰减, 拦截 ``实心块 / 空盒'' 类病态吸引子。\textbf{v2.1 权重}: 5 个拓扑子度量加权几何聚合采用 $(0.20, 0.15, 0.20, 0.30, 0.15)$ (对应 $M_b, M_s, M_j, M_c, M_{\text{bnd}}$), 是经多轮实验、结合人眼评判迭代调优的结果, 使 ``分数排名与视觉评判一致''; $M_c$ 改为纯线性 (旧版 $/0.8$ 截断使 $M_c \in [0.8, 1.0]$ 全部饱和至 1.0, 丧失了区分度). 旧版权重下 ES 排序前列的 4 个真迷宫 ($M_c \in [0.74, 0.98]$) 几乎无区分度; 新权重下, 视觉上更接近迷宫形态的 (b) chebyshev-1 ($M_c = 0.98$) 升至榜首, 视觉上走廊感较弱的 (a) manhattan-2 ($M_c = 0.74$) 调整至第三, 实现了 ``分数排名与视觉评判一致'' 的关键修复。

\noindent\textbf{15-pattern 对照基准.} 6 个真迷宫生成器 (DFS, Kruskal, Prim, Growing Tree, Sidewinder, Binary Tree) 与 9 个几何反例 (Spiral, Fractal Tree, Horizontal Stripes, Random Noise 50\%, Random Noise 30\%, Checkerboard, Diagonal Stripes, Concentric Rings, Honeycomb) 构成基准数据集。$maze\_quality$ 在真迷宫上均值 0.711, 在伪迷宫上均值 0.000, 间隔 +0.711, 误判 0/9. 同一数据集上, Bellot 2021 $F = \nu / \delta$ 的真伪间隔为 $-191.3$, 方向反转, 9 个伪迷宫中有 4 个 (3 个 $F = 0$ 的图案 Horizontal Stripes / Concentric Rings / Honeycomb, 以及 $F = 11.85$ 的 Spiral) 排名低于真迷宫最低值 ($F = 15.33$, DFS), 即 Bellot $F$ 错误地将 Spiral 判为 ``比 DFS 更像迷宫''.

\noindent\textbf{两个尺度的演化扫描.} 小范围扫描: 128 次全交叉运行 (8 mask $\times$ 4 maxFam $\times$ 4 种子, $40 \times 60$ 网格, pop = 200, gens = 500), 其中 122 次成功, 最高分 0.8233 (manhattan-2, maxFam=8, seed 444). 大范围扫描: 5 次运行 (仅 manhattan-2 / maxFam=8 配置, $500 \times 2000$ 网格, pop = 500, gens = 2000, 5 个随机种子), 全部完成, 最高分 0.8195 (seed 444), 5-run 均值 0.8020. 两次扫描峰值相差 0.0037, 表明 $40 \times 60$ 配置已接近该配置的理论性能上界。

\noindent\textbf{两类瓶颈机制及其区分.} (1) 预算受限型 (chebyshev-4, 0.157): $9 \times 9$ 邻域对应的 $2^{103}$ 单族搜索空间与 $10^5$ 预算形成高维无梯度平台, 通过提高预算或缩小 mask 维度可缓解; (2) 表征受限型 (manhattan-1, 0.421): 4 邻域无法稳定涌现走廊结构, 必须更换模板。该二分类本身是本文的方法论贡献, 能够避免在错误的瓶颈方向上投入调优资源。

\vspace{0.5em}
\begin{center}
{\large\bfseries\textcolor{Vermilion}{Abstract}}
\end{center}

\noindent
This paper addresses the problem of determining whether a binary grid produced by cellular automaton (CA) evolution constitutes a maze. Conway-style CA evolution tends to produce attractors that ``look maze-like but are not'' (solid blocks, empty boxes, regular stripes, random noise, etc.), so the question demands a quantitative answer. The paper makes two principal contributions.

\noindent\textbf{FamilyMask} generalises Conway B/S rules to a multi-family chromosome representation. Up to 16 families may run in parallel; each family owns its $9 \times 9$ neighbourhood mask (the centred footprint excluding the centre, 80 cells), birth set, survival set, and 4-bit priority. At each CA step, the active families are scanned in priority order; the first match determines the next state. With one active family the rule space reduces to the standard B/S string ($2^{18}$); with $k$ active families the space expands to $2^{103}$ per family and $2^{1648}$ for the full 16-slot chromosome.

\noindent\textbf{maze\_quality} is an eight-dimensional geometric score. Five topology sub-metrics (corridor dominance $M_b$, longest-path share $M_s$, junction share $M_j$, largest-component share $M_c$, outer-frame suppression $M_{\text{bnd}}$) are combined into $M_{\text{topology}}$ via weighted geometric mean. Three diversity sub-metrics ($2 \times 2$ patch entropy $M_p$, mirror asymmetry $M_a$, $2 \times 2$ adjacency-pair entropy $M_t$) form $M_{\text{diversity}}$. The two sides are combined with $\min$ (rather than the geometric mean), then multiplied by a triangular wall-ratio gate $M_{\text{wall\_gate}}$ on $[0.40, 0.60]$ that zeroes out solid blocks and empty boxes. All eight sub-metrics are continuous in $[0, 1]$ and may therefore be optimised directly by evolution strategies. The $\min$ aggregation is what blocks the failure mode in which one side is high and the other is low. \textbf{v2.1 weights}: the five topology weights are $(0.20, 0.15, 0.20, 0.30, 0.15)$ for $(M_b, M_s, M_j, M_c, M_{\text{bnd}})$, refined through multiple experimental rounds using human visual judgment as the ground truth; $M_c$ is now linear (a previous $/0.8$ clip saturated $M_c \in [0.8, 1.0]$ to 1.0, killing the dimension's gradient). The new weights restore the visual-consistency invariant: chebyshev-1 (visually the most maze-like, $M_c = 0.98$) takes the top rank, while manhattan-2 ($M_c = 0.74$) drops to third, aligning the score ranking with the authors' visual assessment.

\noindent\textbf{15-pattern benchmark.} Six classical maze generators (DFS, Kruskal, Prim, Growing Tree, Sidewinder, Binary Tree) and nine non-maze attractors (Spiral, Fractal Tree, Horizontal Stripes, Random Noise at 30\% and 50\%, Checkerboard, Diagonal Stripes, Concentric Rings, Honeycomb) constitute the benchmark dataset. $maze\_quality$ attains a mean of 0.711 on the true mazes and 0.000 on the pseudo-mazes, with zero misclassifications among the nine pseudo-patterns (gap +0.711). The Bellot 2021 $F = \nu / \delta$ score on the same dataset exhibits a gap of $-191.3$ with the wrong sign (``lower is more maze-like''), and four of nine pseudo-mazes (Spiral at $F = 11.85$ and three $F = 0$ patterns) rank below the genuine DFS ($F = 15.33$). The direction reversal is the smoking gun that motivates $maze\_quality$.

\noindent\textbf{Two scales of evolution sweeps.} Small sweep: 128 full-factorial runs (8 masks $\times$ 4 maxFam $\times$ 4 seeds, $40 \times 60$ grid, pop = 200, gens = 500), 122 successful, top score 0.8233 (manhattan-2, maxFam = 8, seed 444). Large sweep: 5 runs (manhattan-2 / maxFam = 8 only, $500 \times 2000$ grid, pop = 500, gens = 2000, 5 random seeds), all 5 complete, top 0.8195 (seed 444), 5-run mean 0.8020. The two top scores differ by 0.0037, indicating that the $40 \times 60$ configuration is already near the configuration's theoretical ceiling.

\noindent\textbf{Two failure modes and a clean diagnostic that distinguishes them.} chebyshev-4 reaches a mean of 0.157 because its $9 \times 9$ neighbourhood induces a $2^{103}$ per-family space, and $10^5$ evaluations are too sparse to escape the gradientless plateau (budget-limited failure, solvable by raising the budget or shrinking the mask). manhattan-1 reaches 0.421 because the 4-neighbour topology cannot sustain the second-order survival conditions that corridors require (representation-limited failure, requires a different template). The clear separation of these two categories prevents wasted tuning effort in the wrong direction and is itself one of the paper's methodological contributions.

\noindent All code, data, figures, and the 15-pattern benchmark dataset are released as open source.
